---
title: "Semmelweis&Handwashing"
author: Raka Adinugraha
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse) 
library(lubridate)
```

# Dr. Ignaz Semmelweis

Born in 1818, [Dr. Ignaz Semmelweis](https://en.wikipedia.org/wiki/Ignaz_Semmelweis) is a Hungarian physician and active at the Vienna General Hospital during 1840s. Now known as the pioneer of antispetic procedures.

At that time, giving a birth was very dangerous because of [childbed fever](https://en.wikipedia.org/wiki/Puerperal_fever). A deadly disease affecting a woman that just has given birth. In the early 1840s, as many as 10% of the women giving birth die from it in the Vienna general hospital. Dr. Semmelweis knows the cause: it's the contaminated hands of the doctors delivering the babies. Instead, they ridiculed Dr. Semmelweis finding and ignoring him. 

In this project, we going to reanalyze the data of Dr. Semmelweis's finding on the importance of *handwashing*. First, let's start by looking at the data that shows there is something wrong at the Vienna General Hospital that boggles Dr. Semmelweis at that time.

```{r }
# Yearly data by clinic
Yearly_deaths <- read_csv("yearly_deaths_by_clinic.csv", col_types = cols(
  births = col_number(),
  deaths = col_number()))
knitr::kable(Yearly_deaths)
```

# The alarming number of deaths

From the table above, we can see a relatively high number of woman died as a result of giving childbirth at two maternity clinic in Vienna General Hospital. We can notice that it was indeed dangerous to giving births at that time. Most of them died from childbed fever. Now let's add *proportion of deaths* out of woman giving a birth to the table.

```{r }
Yearly_deaths %>%
  mutate(proportion_deaths = deaths / births) -> Yearly_deaths
knitr::kable(Yearly_deaths)
```

And now lets plot the proportion of deaths of both maternity clinic to see if we can get additional information. 

```{r }
Yearly_deaths %>%
  ggplot(aes(x = year, y = proportion_deaths, color = clinic)) +
  geom_line(size = 1) +
  labs(x = "Year", y = "Proportion Deaths") +
  theme_minimal() +
  theme(legend.title = element_blank())
```

As we can see, the deaths on the first clinic always averagely higher than on the second clinic, about 10% average mortality rate compared to less than 4% on the second clinic.

Both maternity clinics used almost the same techniques in delivering babies, but the fact that the first clinic had a much higher mortality rate puzzled Dr. Semmelweis.

There is only one major difference between the two clinics. The first clinic was the teaching service for medical students, while the second clinic was for the instruction of midwives only since 1841. While the midwives only tended to woman giving birth, the medical students and the doctors also spent their time in autopsy room performing post mortem examination on a human corpse. And this explains why the second clinic saw a much lower mortality rate because the midwives not engaged in autopsies and had no contact with the corpses.

After the meticulous effort, Dr. Semmelweis suspect that something from the corpses somehow makes its way to the woman giving birth from the medical student's hands and causing chilbed fever. In June 1847, Dr. Semmelweis made handwashing mandatory. 

And let's see the monthly data from clinic 1. 

```{r }
# Monthly data from clinic 1
Monthly_deaths <- read_csv("monthly_deaths.csv", col_types = cols(
  date = col_date(format = "")))

# Add proportion of deaths
Monthly_deaths %>%
  mutate(proportion_deaths = deaths / births) -> Monthly_deaths
head(Monthly_deaths)
```

Now let's plot the data from the table above. Births count with a blue dot, and deaths count with a red dot. In the plot below we can already see the effectiveness of handwashing even though we haven't marked it yet. 

```{r }
ggplot(Monthly_deaths) +
  geom_segment(aes(x=date, xend=date, y=deaths, yend=births), color="grey") +
  geom_point(aes(x=date, y =births), color="dodgerblue4") +
  geom_point(aes(x=date, y=deaths), color="darkred") +
  theme_minimal() +
  labs(x = "Year", y = "Deaths & Births Number")
```

# The handwashing started

After the handwashing made mandatory, the mortality rate drastically reduced. We can see it more clearly in the graph below. 

```{r }
# Add TRUE/FALSE to identified month with handwashing obligatory
Monthly_deaths %>%
  mutate(handwashing_started = date >= as_date('1847-06-01')) -> Monthly_deaths

Monthly_deaths %>%
  ggplot(aes(x = date, y = proportion_deaths, color = handwashing_started)) +
  geom_line(size = 1) +
  labs(x = "Year", y = "Proportion Deaths") +
  theme_minimal()
```

## Fewer deaths after handwashing? 

As the graph shows above, it's indeed that handwashing reduces drastically the monthly proportion of deaths. Now let's see how much it's reduced on average.

```{r }
Monthly_deaths %>% 
  group_by(handwashing_started) %>%
  summarise(mean_proportion_deaths = mean(proportion_deaths)) -> monthly_sum
knitr::kable(monthly_sum)
```

It reduces the proportion of deaths by about 8 percentage points. From 10% on average to about 2% when handwashing was enforced. And by using t-test, we could look at confidence interval to get a feeling for the uncertainty around how much handwashing reduces mortalities rate. 

```{r }
# Using t-test to calculate 95% confidence interval
t_result <- t.test(proportion_deaths ~ handwashing_started, data = Monthly_deaths)
t_result
```

# After the handwashing

When the doctors and medical students didn't wash their hands, it increased the proportion of deaths by around 6.7 to 10 percentage points, according to a 95% confidence interval from the previous section. 

Despite this solid evidence of this finding, the contemporary scientist instead ridiculed Dr. Semmelweis's theory - that some "substance"(what we know today as *bacteria*) from autopsy room corpses caused the childbed fever. The medical community largely rejected his theory and finally, at 1849 Dr. Semmelweis forced to leave the Vienna General Hospital for good.

One reason for this was that statistics and statistical arguments were uncommon in medical science in the 1800s. If he had access to the analysis we've just put together he might haven't got a rejection from the contemporary scientist and the medical community. 

The [germ theory of disease](https://en.wikipedia.org/wiki/Germ_theory_of_disease) that offers a theoretical explanation of Dr. Semmelweis  findings only get developed decades later after Dr. Semmelweis's deaths. Only after this, his practice gets widespread acceptance and then he is considered a pioneer of antiseptic procedure. 

Up to this day, it still true to some extent that handwashing could save lives. 